---
title: Summary Output
author: Sea State
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    fig_caption: true
    toc: true
    toc_float: true
    code_folding: show
---


```{r,message=FALSE,warning=TRUE}
	.BASE <- "../models_2015/sitka/ham"
	library(ggplot2)
	library(dplyr)
	library(tidyr)
	library(pander)
	# Read in the data from the model report, par, and cor files.
	source("./globals.R")
	source("ham.R")
	D <- read.admb(.BASE)
```



# Input Data
The following is a summary of the input data provided in the data file.

## Model dimensions
The first six elements in the following code snippet shows the first and last years of data (**dat_syr, dat_nyr**), the first and last years of the population dynamics model (**mod_syr, mod_nyr**), and the youngest age-class (**sage**) and the oldest age class (**nage**). Note that the oldest age-class is modelled as a plus group.

```{r,message=FALSE,warning=FALSE}
	# source("ham.R")
	pander(D[1:6])
```

## Total Annual Catch

```{r,message=FALSE,warning=FALSE}
	colnames(D[['data_ct_raw']]) <- c("Year","Catch","log.se")
	plot.catch(D,y="Catch (short tons)")
	panderOptions("digits",3)
	panderOptions("keep.trailing.zeros",TRUE)

	pander(as.data.frame(D$data_ct_raw))
```

## Empirical Weight-at-age
The weight-at-age data have the units of grams.  There are two input weight-at-age matrixes, the first is for the observed spawning weight-at-age, and the second is the observed commercial catch weight-at-age.  The row dimensions of each matrix are from dat_syr -- dat_nyr.
```{r,message=FALSE,warning=FALSE}
	colnames(D[['data_sp_waa']]) <- c("Year",paste(D$sage:D$nage))
	plot.waa(D)

	pander(D$data_sp_waa)
```





## Catch Composition
The next two blocks of data input correspond to the Commercial age composition and spawn--sample age composition data, respectively.  A row must be specified for each year, and if there are no data for a given year, the entry is simply -9.  For example, see the missing data in 1977 in the example below. 

```{r,message=FALSE,warning=FALSE,fig.cap="Commercial age-proportions where the area of each circle is proportional to the proportion-at-age in a given year and the colors are unique to each cohort.",fig.height=3}
	plot.comp(D,nm = "data_cm_comp")
```
```{r,message=FALSE,warning=FALSE}
	colnames(D[['data_cm_comp']]) <- c("Year",paste(D$sage:D$nage))
	pander(D$data_cm_comp)
```

## Spawn Composition
```{r,message=FALSE,warning=FALSE,fig.cap="Spawn-sample age-proportions where the area of each circle is proportional to the proportion-at-age in a given year and the colors are unique to each cohort.",fig.height=3}
	plot.comp(D,nm = "data_sp_comp")
```
```{r,message=FALSE,warning=FALSE}
	colnames(D[['data_sp_comp']]) <- c("Year",paste(D$sage:D$nage))
	panderOptions("digits",6)
	panderOptions("keep.trailing.zeros",TRUE)
	pander(D$data_sp_comp)
```


## Egg Deposition Data
The egg deposition data follows the spawn--sample age composition data. This is a 3 column matrix with row dimensions specified for each *data year*. The index value for years with no survey data are set equal to -9.  

```{r,message=FALSE,warning=FALSE}
	plot.datafit(D,fit=FALSE)
```

```{r,message=FALSE,warning=FALSE}
		colnames(D[['data_egg_dep']]) <- c("Year","Index","log.se")
		pander(D$data_egg_dep)

```


## Mile-Days of Milt
The last of the fisheries independent data is just before the *end of file marker* (#EOF 999).  This data structure has the same dimensions as the previous table.  There must be a row for each *data year*, and missing data is substituded with a -9 for the index value.
```{r,message=FALSE,warning=FALSE}
		colnames(D[['data_mileday']]) <- c("Year","Index","log.se")
		(D$data_mileday)

```

----

# Model Output (MLE)

The following output is based on the Maximum likelihood estimates contained in the parameter and report files.

## Summary statistics
```{r,message=FALSE,warning=FALSE}
	pander(D$fit[1:5])
```
## Parameter estimates
```{r,message=FALSE,warning=FALSE}
	df <- data.frame(Parameter=as.character(D$fit$names),
	                 Estimate =round(D$fit$est,3),
	                 StdDev=round(D$fit$std,3)) %>%
				tbl_df()
	pander(as.data.frame(df))
```

## Residuals
```{r,message=FALSE,warning=FALSE}
	plot.resd(D,nm="resd_cm_comp")
	plot.resd(D,nm="resd_sp_comp")
```









